### 2021-07-13 ch/1-introduction.tex
Lot's of changes

\section{Relations to other fields and clarification of terms}
\subsection{aka: Isn't this thesis actually about XY instead of fairness?}
The thesis is mostly written from the perspective of algorithmic fairness,
but it touches on other fields as well, like \emph{domain shift} and \emph{causality}.
Furthermore, the thesis makes use of concepts like \emph{transferable representations}
and \emph{interpretability}.
I summarise the areas here and discuss their relevance to the presented work.

>>>>

\section{Relations to other fields and clarification of terms}
\subsection{aka: Isn't this thesis actually about XY instead of fairness?}
The thesis is principally written from the perspective of algorithmic fairness,
but it touches on other fields as well, like \emph{domain shift} and \emph{causality}.
Furthermore, the thesis makes use of concepts like \emph{transferable representations}
and \emph{interpretability}.
I summarise the areas here and discuss their relevance to the presented work.

### 2021-07-13 ch/2-related-work.tex
Lot's of changes

\chapter{Related work}\label{ch:related-work}
%
% =========================
%      old introduction
% =========================
% \section{Introduction}\label{introduction}
% Fairness in Machine Learning is a topic that has recently received increasing attention.
% Machine Learning has become very widespread in our society
% which brings some dangers with it.
% Decisions such as credit eligibility and hiring are increasingly done automatically.
% However, the algorithms that are used in these decisions can have strong biases
% that stem from the training data.
% Often, the algorithms are even more biased than the training data.
% \citet{kamishima2012fairness} give a great illustration for this
% (adapted from \citet{calders2010three}).
% The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
% which contains records of individuals with thirteen attributes, among them age and gender.
% The task is to predict whether or not the individual earns more than \$50K per year.
% In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
% but of the women only 11\% have a positive label.
% With this ratio of about 3:1, the dataset is already skewed.
% However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Na\"ive Bayes classifier} on the data,
% the ratio in the predictions is even more unfair:
% the amount of women who get a positive prediction is even lower than 11\%.
% If the Na\"ive Bayes classifier does not get access to the gender information,
% the predictions will be a bit fairer,
% but they are still more unfair than the original dataset labels.
%
% This phenomenon of predictions being more unfair than an already skewed dataset
% is due to the generalisation that classifiers have to make.
% It is easier to classify all but the most outstanding women as low income
% than to learn a complicated decision boundary.
% Furthermore, removing the gender information does not solve the problem
% because the gender can be inferred from other data.
%
% The following is a review of the literature on Fairness in Machine Learning.
% Several different approaches to and definition of Fairness are presented.
% The focus is on identifying the major trends in the field from the beginning.
\section{Two views of the dataset bias problem}\label{two-views-of-the-dataset-bias-problem}
As alluded to in the introduction,
there is a divide in the relevant literature in how the problem of dataset bias is approached.
On the one hand, there is the ``ground-truth-centric'' view -- expressed through most of the introduction --
that the training set is biased but the deployment setting is not.
By \emph{deployment setting} I am here referring to the real distribution
that will be encountered by the machine learning system once it is in use.

@@

>>>>

\chapter{Related work}\label{ch:related-work}
%
% =========================
%      old introduction
% =========================
% \section{Introduction}\label{introduction}
% Fairness in Machine Learning is a topic that has recently received increasing attention.
% Machine Learning has become very widespread in our society
% which brings some dangers with it.
% Decisions such as credit eligibility and hiring are increasingly done automatically.
% However, the algorithms that are used in these decisions can have strong biases
% that stem from the training data.
% Often, the algorithms are even more biased than the training data.
% \citet{kamishima2012fairness} give a great illustration for this
% (adapted from \citet{calders2010three}).
% The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
% which contains records of individuals with thirteen attributes, among them age and gender.
% The task is to predict whether or not the individual earns more than \$50K per year.
% In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
% but of the women only 11\% have a positive label.
% With this ratio of about 3:1, the dataset is already skewed.
% However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Na\"ive Bayes classifier} on the data,
% the ratio in the predictions is even more unfair:
% the amount of women who get a positive prediction is even lower than 11\%.
% If the Na\"ive Bayes classifier does not get access to the gender information,
% the predictions will be a bit fairer,
% but they are still more unfair than the original dataset labels.
%
% This phenomenon of predictions being more unfair than an already skewed dataset
% is due to the generalisation that classifiers have to make.
% It is easier to classify all but the most outstanding women as low income
% than to learn a complicated decision boundary.
% Furthermore, removing the gender information does not solve the problem
% because the gender can be inferred from other data.
%
% The following is a review of the literature on Fairness in Machine Learning.
% Several different approaches to and definition of Fairness are presented.
% The focus is on identifying the major trends in the field from the beginning.
%
\begin{figure}[tbp]
  \resizebox{\textwidth}{!}{
    \begin{forest}
    forked edges,
    for tree={rounded corners, edge+={darkgray, line width=1pt}, draw=darkgray, align=center, anchor=children},
    before packing={where n children=3{calign child=2, calign=child edge}{}},
    before typesetting nodes={where content={}{coordinate}{}},
    where level<=1{line width=2pt}{line width=1pt},
    [methods for algorithmic fairness
      [data de-biasing
        [dropping\\features]
        [fair repre-\\sentations
          [non-parametric
            [HSIC]
            [MMD]
          ]
          [adversarial\\networks
            [sample-wise, draw=blue]
            [set-wise, draw=blue]
          ]
        ]
      ]
      [post-hoc methods, line width = 2pt
        [separate\\models]
        [threshold\\shifting]
      ]
      [
        [change to train-\\ing procedure, line width = 2pt
          [modified\\sampling]
          [
            [change to loss
              [change of\\optimisation target, draw=blue]
              [constrained\\optimisation]
            ]
          ]
          [loss\\reweighting]
        ]
      ]
    ]
    \end{forest}
  }
  \caption{%
    Taxonomy of fairness methods.
    The three categories with blue borders are those categories
    to which the proposed models in this thesis belong (see \chapref{ch:content} for details).
    This diagram does not show all the axes along which methods can vary.
    In particular, it does not distinguish between different intended \emph{settings}.
  }%
  \label{fig:fairness-taxonomy}
\end{figure}%
%
\section{Two views of the dataset bias problem}\label{two-views-of-the-dataset-bias-problem}
As alluded to in the introduction,
there is a divide in the relevant literature in how the problem of dataset bias is approached.
On the one hand, there is the ``ground-truth-centric'' view -- expressed through most of the introduction --
that the training set is biased but the deployment setting is not.
By \emph{deployment setting} I am here referring to the real distribution
that will be encountered by the machine learning system once it is in use.

@@

\Figref{fig:fairness-taxonomy} provides a high-level overview
of one possible categorisation of the methods presented here.
It is meant as a rough guide through the chapter.

### 2021-07-13 ch/3-structure.tex
Lot's of changes

@@

@@

The method implementing this general strategy, and presented in \citet{kehrenberg2020nullsampling} (\chapref{ch:paper2}),
is based on the idea of \emph{null-sampling},
which refers to zeroing out part of an encoding,
and then reconstructing the modified encoding as if it were a normal encoding.
In order to apply null-sampling, an encoding of the input \(x\) is learned that is split into two parts:
\(z_u\), which has no information about the spurious attribute \(s\),
and \(z_b\), which has all the remaining information needed to reconstruct \(x\) that is not contained in \(z_u\).
\(z_u\) is ensured to be not predictive of \(s\) via adversarial training.
During null-sampling, \(z_b\) is zeroed out, and after decoding it,
we obtain an invariant representation \emph{in the data domain}.
The fact that it is in the data domain makes it interpretable or inspectable.

The described method works particularly well with \acp{INN},
as they ensure that no information is lost that is unrelated to \(s\).

We perform experiments on the Coloured MNIST dataset (as described in \secref{sec:groundtruth-centric-view-of-bias}),
which has a one-to-one mapping between the class label (\ie, digit) and the spurious attribute (colour)
in the training set.
As colour is ``easier'' to learn, a neural network will learn to predict \(s\) instead of \(y\).
For additional experiments on the CelebA dataset and the UCI Adult Income dataset,
we deliberately apply sampling bias to the training set and then apply our method.

@@

% \section{Claims and contributions}%
\section{List of publications and author contributions}%
\label{sec:claims-contributions}
This thesis is based on 3 publications, corresponding to \rangechapref{ch:paper1}{ch:paper3}.
The first one is concerned with \emph{label bias} and the other two with \emph{sampling bias}.

>>>>

% - suddenly images?
% - (explain in the linking chapter)
% - is it specific to images?
% - explain how it works for tabular data

@@

The method is validated with experiments
on the UCI Adult Income dataset and the ProPublica/COMPAS dataset,
which have been mentioned several times in \chapref{ch:related-work},
and which are the most common tabular fairness datasets.
While both these datasets comprise only tabular data,
there is nothing in principle that stops this method from being used for other kinds of data.
The choice to use these datasets and not others
was predominantly made for easier comparison to baselines.

@@

The method implementing this general strategy, and presented in \citet{kehrenberg2020nullsampling} (\chapref{ch:paper2}),
is based on the idea of \emph{null-sampling},
which refers to zeroing out part of an encoding,
and then reconstructing the modified encoding as if it were a normal encoding.
In order to apply null-sampling, an encoding of the input \(x\) is learned that is split into two parts:
\(z_u\), which has no information about the spurious attribute \(s\),
and \(z_b\), which has all the remaining information needed to reconstruct \(x\) that is not contained in \(z_u\).
\(z_u\) is ensured to be not predictive of \(s\) via adversarial training.
During null-sampling, \(z_b\) is zeroed out, and after decoding it,
we obtain an invariant representation \emph{in the data domain}.
The fact that it is in the data domain makes it interpretable (or inspectable)
as defined in \chapref{ch:introduction}.

The described method works particularly well with \acp{INN},
as they ensure that no information is lost that is unrelated to \(s\).
However, the price to pay for using \acp{INN} is higher memory requirement and slower training.
Thus, we also present a variant of the method using \iac{VAE},
which does not have the guarantee about preserving information,
but also does not suffer from the increased training cost as much.
\acp{VAE} are similar to \acp{INN}
in that their encoding conforms to a specific probability distribution,
from which we can sample our null-samples.
The choice between the two presented variants comes down
to whether the user is willing to trade higher training costs for higher-quality results.
However, the key element is any kind encoder -- producing a split-encoding --
whose output can be subjected to adversarial training,
and other kinds of encoders will potentially work as well.

We perform experiments on the Coloured MNIST dataset (as described in \secref{sec:groundtruth-centric-view-of-bias}),
which has a one-to-one mapping between the class label (\ie, digit) and the spurious attribute (colour)
in the training set.
As colour is ``easier'' to learn, a neural network will learn to predict \(s\) instead of \(y\).
For additional experiments on the CelebA dataset (another image dataset)
and the UCI Adult Income dataset (a tabular dataset),
we deliberately apply sampling bias to the training set and then apply our method.
For the tabular dataset, an autoencoder is trained to produce a continuous representation,
which is then fed into the \ac{INN} or \ac{VAE} model.
Thus, we demonstrate that the method is \emph{general}
in the sense that it is applicable to both image-based and tabular datasets
and furthermore should be applicable to other modalities as well.
For the main experiments, we focus on image datasets,
because they are easiest to visualise in a document.

@@

Experiments are performed on the same datasets as in the previous papers:
Coloured MNIST, UCI Adult Income and CelebA.
Again, these datasets were chosen, because image data is easy to visualise,
and because demonstrating the method on tabular data provides evidence for the generality.

% \section{Claims and contributions}%
\section{List of publications and author contributions}%
\label{sec:claims-contributions}
This thesis is based on 3 publications, corresponding to \rangechapref{ch:paper1}{ch:paper3}.
% The first one is concerned with \emph{label bias} and the other two with \emph{sampling bias}.

### 2021-07-13 paper2/4-experiments.tex
Lot's of changes

\begin{figure}[tb]
    \centering
    % \includegraphics[width=0.9\textwidth]{./results/cmnist/cmnist_results.pdf}
    % \includegraphics[width=0.9\textwidth]{paper2/Figures/cmnist_new.pdf}
    \includegraphics[width=0.7\textwidth]{paper2/Figures/cmnist_new_no_hgr.pdf}
    \caption{
        % Accuracy and correlation (HGR) of predictions $\hat{y}$ and $s$ of our approach in comparison with other baseline models on the cMNIST dataset, for different standard deviations ($\sigma$) for the colour sampling.
        Accuracy of our approach in comparison with other baseline models on the cMNIST dataset, for different standard deviations ($\sigma$) for the colour sampling.
    }%
    \label{fig:cmnist_chart}
\end{figure}
\subsection{Experimental results}
% \noindent Our model can be applied to tabular data and image data alike.
% Due to space constraints, the experiments with the tabular dataset \emph{UCI Adult} are situated in the Appendix.
%To illustrate the effectiveness and generality of our framework, we apply our models to a tabular dataset, in the UCI Adult dataset,
We report the results from two image datasets.
cMNIST, a synthetic dataset, is a good starting point for evaluating our model due to the direct control we have over the biasing.
CelebA, on the other hand, is a more practical and challenging example.
% To illustrate the generality of our approach,
We also test our method on a tabular dataset, the Adult dataset.
%For full details regarding the architectures of our models and their optimisation, see the Appendix.

@@

Inspection of the null-samples shows that both the cVAE and cFlow model succeed in removing almost all colour information, which is supported quantitatively by fig.~\ref{fig:cmnist_chart}.
While the cVAE outperforms cFlow marginally at low $\sigma$ values, performance degrades
%rapidly
as this increases.
This highlights the problems with the conditional decoder we anticipated in section~\ref{conddec}.
The lower $\sigma$, and therefore the variation in sampled colour, is, the more reliably the $s$ label, corresponding to the mean of RGB distribution, encodes information about the colour.
For higher $\sigma$ values, the sampled colours can deviate far from the mean and so the encoder must incorporate information about $s$ into its representation if it is to minimise the reconstruction loss.
cFlow, on the other hand, is consistent across $\sigma$ values.

@@

We also assess performance for different mixing factors ($\eta$) which correspond to varying degrees of bias in the training set
(see fig.~\ref{fig:celeba-multiplot}).
This is to verify that the model does not \emph{harm} performance when there is not much bias in the training set.
For these experiments, the model is trained once on the representative set and is then used to encode different training sets.
The results show that for the intermediate values of $\eta$, our model incurs a small penalty in terms of accuracy,
but at the same time makes the results \emph{fairer} (corresponding to an accuracy-fairness trade-off). Qualitative results can be found in fig.~\ref{fig:celeba_cflow} (images from cVAE can be found in section~\ref{sec:qual-results-celeba}).

>>>>

\begin{figure}[tb]
    \centering
    % \includegraphics[width=0.9\textwidth]{./results/cmnist/cmnist_results.pdf}
    % \includegraphics[width=0.9\textwidth]{paper2/Figures/cmnist_new.pdf}
    \includegraphics[width=0.7\textwidth]{paper2/Figures/cmnist_new_no_hgr.pdf}
    \caption{
        % Accuracy and correlation (HGR) of predictions $\hat{y}$ and $s$ of our approach in comparison with other baseline models on the cMNIST dataset, for different standard deviations ($\sigma$) for the colour sampling.
        Accuracy of our approach in comparison with other baseline models on the cMNIST dataset, for different standard deviations ($\sigma$) for the colour sampling.
    }%
    \label{fig:cmnist_chart}
\end{figure}

\begin{figure*}[!htb]
    \centering
    \subfloat[Samples from the cMNIST training set, $\sigma=0$.]{%
        \scalebox{0.3}{\includegraphics[width=\textwidth]{paper2/Images/cmnist/cflow_original_task_x_scale_0.png}}%
        \label{fig:cflow_cmnist_task_train}
    }
    % \hfill
    % \subfloat[Samples from the cMNIST test set with $\sigma=0.02$.]{%
    %     \scalebox{0.225}{\includegraphics[width=\textwidth]{./Images/cmnist/cflow_pretrain_x.png}}%
    %     \label{fig:cflow_cmnist_pretrain}
    % }
    \hfill
    \subfloat[$x_u$ null-samples from the cFlow model.]{%
        \scalebox{0.3}{\includegraphics[width=\textwidth]{paper2/Images/cmnist/cflow_task_xy_scale_0.png}}%
        \label{fig:cflow_cmnist_y}
    }
    \hfill
    \subfloat[$x_b$ null-samples from the cFlow model.]{%
        \scalebox{0.3}{\includegraphics[width=\textwidth]{paper2/Images/cmnist/cflow_task_xs_scale_0.png}}%
        \label{fig:cflow_cmnist_s}
    }
    \caption{
        Sample images from the coloured MNIST dataset problem with $10$ predefined mean colours.
        (a): Images from the spuriously correlated subpopulation where colour is a reliable signal of the digit class-label.
        (b-c): Results of running our approach realised with cFlow on the cMNIST dataset.
        The model learns to retain the shape of the digit shape while removing the relationship with colour.
        A downstream classifier is now less prone to exploiting correlations between colour and the digit label class.
    }\label{fig:cmnist}
\end{figure*}

\subsection{Experimental results}
% \noindent Our model can be applied to tabular data and image data alike.
% Due to space constraints, the experiments with the tabular dataset \emph{UCI Adult} are situated in the Appendix.
%To illustrate the effectiveness and generality of our framework, we apply our models to a tabular dataset, in the UCI Adult dataset,
We report the results from two image datasets.
cMNIST, a synthetic dataset, is a good starting point for evaluating our model due to the direct control we have over the biasing.
CelebA, on the other hand, is a more practical and challenging example.
% To illustrate the generality of our approach,
We also test our method on a tabular dataset, the Adult dataset.
%For full details regarding the architectures of our models and their optimisation, see the Appendix.

@@

Inspection of the null-samples shows that both the cVAE and cFlow model succeed in removing almost all colour information, which is supported quantitatively by fig.~\ref{fig:cmnist_chart},
and qualitatively by fig.~\ref{fig:cmnist}.
While the cVAE outperforms cFlow marginally at low $\sigma$ values, performance degrades
%rapidly
as this increases.
This highlights the problems with the conditional decoder we anticipated in section~\ref{conddec}.
The lower $\sigma$, and therefore the variation in sampled colour, is, the more reliably the $s$ label, corresponding to the mean of RGB distribution, encodes information about the colour.
For higher $\sigma$ values, the sampled colours can deviate far from the mean and so the encoder must incorporate information about $s$ into its representation if it is to minimise the reconstruction loss.
cFlow, on the other hand, is consistent across $\sigma$ values.

@@

\begin{figure*}[tb]
  \centering
  \subfloat[Original images.]{%
      \scalebox{0.3}{\includegraphics[width=\textwidth]{paper2/Images/celeba/train_original_x_2.png}}%
      \label{fig:cflow_celeba_original_x}
  }
  \hfill
  \subfloat[$\bm{x}_u$ null-samples from the cFlow model.]{%
      \scalebox{0.3}{\includegraphics[width=\textwidth]{paper2/Images/celeba/train_reconstruction_y_2.png}}%
      \label{fig:cflow_celeba_recon_y}
  }
  \hfill
  \subfloat[$\bm{x}_b$ null-samples from the cFlow model.]{%
      \scalebox{0.3}{\includegraphics[width=\textwidth]{paper2/Images/celeba/train_reconstruction_s_2.png}}%
      \label{fig:cflow_celeba_recon_s}
  }
  \caption{
    CelebA null-samples learned by our cFlow model, with gender as the sensitive attribute.
    (a) The original, untransformed samples from the CelebA dataset
    (b) Reconstructions using only information unrelated to $s$.
    (c) Reconstruction using only information related to $\neg s$.
    The model learns to disentangle gender from the non-gender related information.
    % Attributes such as \emph{makeup} and \emph{hair length} are also often modified in the process due to inherent correlations between them and the sensitive attribute, which the intepretability of our representations allows us to easily identify.
    Note that some attributes like skin tone seem to change along with gender due to the correlation between the attributes.
    This is especially visible in images (1,1) and (3,2). Only because our representations are produced in the data-domain can we easily spot such instances of entanglement.
  }%
  \label{fig:celeba_cflow}
\end{figure*}
%
We also assess performance for different mixing factors ($\eta$) which correspond to varying degrees of bias in the training set
(see fig.~\ref{fig:celeba-multiplot}).
This is to verify that the model does not \emph{harm} performance when there is not much bias in the training set.
For these experiments, the model is trained once on the representative set and is then used to encode different training sets.
The results show that for the intermediate values of $\eta$, our model incurs a small penalty in terms of accuracy,
but at the same time makes the results \emph{fairer} (corresponding to an accuracy-fairness trade-off). Qualitative results can be found in fig.~\ref{fig:celeba_cflow} (images from cVAE can be found in section~\ref{sec:qual-results-celeba}).
